Highly Available Transactions: Virtues and Limitations\\(Extended Version)

### 2013-07-22 writeup/hat-analysis.tex
Starting to address TA -> RA

\textbf{Transactional atomicity (TA)} is core to ACID
guarantees. Although, at least by the ACID acronym, it is not an
``isolation'' property, TA restricts transactions' ability to view the
effects of partially completed transactions. Under TA, once some of
the effects of a transaction $T_i$ are observed by another transaction
$T_j$, thereafter, all effects of $T_i$ are observed by
$T_j$. Together with item cut isolation, TA prevents Read Skew
anomalies (Berenson et al.'s A5A) and is useful in several contexts
such as maintaining foreign key constraints, consistent global
secondary indexing, and maintenance of derived data. In the example
below, under TA, because $T_2$ has read $T_1$'s write to $y$, $T_2$
must observe $b=c=1$ (or later versions for each key):
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$, $a=1$, or a later version of
$x$. Notably, TA requires disallows reading intermediate writes
(Adya's $G1b$): observing all effects of a transaction implicitly
requires observing the final (committed) effects of the transaction as
well.

Perplexingly, discussions of TA are absent from existing treatments of
weak isolation. This is perhaps again due to the single-node context
in which prior work was developed: on a single server (or a fully
replicated database), TA is achievable via lightweight locking and/or
local concurrency control over data items~\cite{gstore,
  kemme-thesis}. In contrast, in a distributed environment, TA over
arbitrary groups of non-colocated items is considerably more difficult
to achieve with high availability: servers need to ensure that,
regardless of replica selection, all of a client's accesses will read
a sufficiently up-to-date/``consistent'' set of data items. However,
replicas do not need to agree on when to choose to reveal sets of
values to clients, which would require consensus. Instead, as we see
below, TA only requires the equivalent of reliable broadcast (with
some additional client metadata) and is achievable in HAT systems.

% TA can be achieved via multi-versioning. If clients attach vector
% clocks to every write (incrementing their own position for each
% transaction) and replicas store all writes via multi-versioning, then
% clients can safely determine which sets of data items to read; this
% approach is adopted by Swift~\cite{swift} and, to a lesser-extent,
% bolt-on causal consistency~\cite{bolton}. These systems also provide
% causal consistency and are subsequently sticky available (each is
% implemented via client-side caching). However, TA is achievable
% without stickiness via an alternate algorithm: if servers wait to
% reveal writes until they are present on all replicas, clients do not
% need to be sticky.

We sketch a highly available TA algorithm and provide greater detail
in Appendix B.
%our extended technical report~\cite{hat-tr}.
We begin with our Read Committed algorithm, but replicas wait to reveal new writes to readers
until all of the replicas for the final writes in the transaction have
received their respective writes (are \textit{pending
  stable}). Accordingly, replicas reveal writes asynchronously;
different replicas may detect pending stability at different times. To
prevent clients from reading incomplete portions of a transaction's
writes as they are being revealed (e.g., a client reads a data item
that from a replica that has detected pending stability, then attempts
to read another item from another replica that has not yet detected
pending stability), clients include additional metadata with each
write: a single timestamp for all writes in the transaction (e.g., as
in Read Uncommitted) and a list of items written to in the
transaction. When a client reads a write, the write's timestamp and
list of items form a lower bound on the versions that the client
should read for the other items. When a client reads any data item, it
attaches a timestamp to its request representing the current lower
bound for that item. Replicas use this timestamp to respond with
either a write matching the timestamp or a pending stable write with a
higher timestamp. Clients may discard any metadata upon transaction
commit, and servers keep two sets of writes for each data item: the
write with the highest timestamp that is pending stable and a set of
writes that are not yet pending stable.

This implementation is entirely master-less and operations never block
due to replica coordination. Moreover, there are several possible
optimizations that, for brevity, we do not describe here. However, one
key property is that we do not dictate when writes become visible;
accordingly, the algorithm is highly available. We can optimize write
visibility via sticky availability by sticking clients are with
disjoint groups of servers: once all of a transaction's writes are
pending stable within a group of servers, writes can be made visible
(as opposed to once they are pending stable with respect to all
servers). For example, in a multi-datacenter setting, all clients and
servers within a datacenter may form a sticky group.

@@

We illustrate the hierarchy of available, sticky available, and
unavailable consistency models we have discussed in
Figure~\ref{fig:hatcompared}. Many models are simultaneously
achievable, but we find several particularly compelling. If we combine
all sticky-HAT guarantees, we have transactional, causal snapshot
reads (i.e., Causal Transactional Predicate Cut Isolation). If we
combine TA and P-CI, we have transactional snapshot reads. We can
achieve RC, MR, and RYW by simply sticking clients to servers. We can
also combine unavailable models---for example, an unavailable system
might provide PRAM and One-Copy
Serializability~\cite{daudjee-session}.

@@

\begin{table}[t!]
\begin{tabular}{| c | p{6cm} | }\hline
HA & Read Uncommitted (RU), Read Committed (RC), Transactional
Atomicity (TA), Item Cut Isolation (I-CI), Predicate Cut Isolation
(P-CI), Writes Follow Reads (WFR), Monotonic Reads (MR), Monotonic
Writes (MW)\\\hline Sticky HA & Read Your Writes (RYW), PRAM,
Causal\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
Snapshot Isolation (SI)\lostupdate, Repeatable Read
(RR)\lostupdate\rwskew, One-Copy Serializability
(1SR)\lostupdate\rwskew, Recency\linearizable, Safe\linearizable,
Regular\linearizable, Linearizability\linearizable, Strong
1SR\lostupdate\rwskew\linearizable \\\hline
\end{tabular}
\caption{Summary of highly available, sticky highly available, and
  unavailable models considered in this paper. Unavailable models are
  labeled by cause of unavailability: preventing lost
  update\lostupdate, preventing write skew\rwskew, and requiring
  recency guarantees\linearizable.}
\label{table:hatcompared}
\end{table}

@@

 \node[draw=none,fill=none] (ici) at (1.2, 0) {I-CI};
 \node[draw=none,fill=none] (pci) at (1.75, 1.2) {P-CI};
 \node[draw=none,fill=none] (rc) at (-1.2, .95) {RC};
 \node[draw=none,fill=none] (ru) at (-1.2, 0) {RU};

 \node[draw=none,fill=none] (ta) at (-.2, 0) {TA};

@@

 \node[noha](recency) at (7.7, 0) {recency};
 \node[noha](safe) at (7.7, 1) {safe};
 \node[noha](regular) at (7.7, 2) {regular};
 \node[noha](linearizable) at (7.7, 3) {linearizable};
 \node at (4.8, 2) [sticky] (causal) {causal};
 \node at (4.8, 1) [sticky] (pram) {PRAM};
 \node[noha] (cs) at (-1.2, 2) {CS};
 \node[noha] (rr) at (-.2, 2.7) {RR};
 \node[noha] (si) at (1.75, 2.3) {SI};
 \node[noha] (1sr) at (1.2, 3.2) {1SR};
 \node[noha] (ssr) at (3.85, 3.6) {Strong-1SR};

@@

 \draw [->] (ru) -- (rc);
 \draw [->] (ici) -- (pci);

@@

 \draw [->, red] (rc) -- (cs);
 \draw [->, red] (cs) -- (rr);
 \draw [->, red] (pci) -- (si);
 \draw [->, red] (ici) -- (rr);
 \draw [->, red] (rr) -- (1sr);
 \draw [->, red] (si) -- (1sr);
 \draw [->, red] (ta) -- (si);
 \draw [->, red] (ta) -- (rr);
 \draw [->, red] (causal) -- (linearizable);
 \draw [->, red] (ryw) -- (safe);

@@

\vspace{.5em}\noindent\textbf{Linearizability and Transactional
  Atomicity} The relationship between linearizability and
transactional atomicity is non-obvious. TA dictates that writes to
multiple keys across multiple servers are made visible to readers all
at once, while linearizability dictates that writes to a single key on
multiple servers are made visible to all readers at once---what is
different? First, in linearizable (and safe and regular) systems,
writes are made visible to clients \textit{immediately} after they
finish. With transactional atomicity, there is no recency
guarantee. Second, in linearizable systems, all clients see all writes
at the same time. With TA as defined here, clients may see writes at
different times depending on which replicas they contact. We are not
aware of an analogous model in the distributed systems
literature. Accordingly, despite apparent similarities, TA is
incomparable with and less coordination intensive (by availability
measures) than linearizability.

>>>>

Atomicity, informally guaranteeing that either all or none of
transactions' effects should succeed, is core to ACID
guarantees. Although, at least by the ACID acronym, it is not an
``isolation'' property, atomicity properties also restrict the updates
visible to other transactions. Accordingly, here, we consider the
\textit{isolation} effects of atomicity, which we call \textbf{Read
  Atomic (RA)} isolation.  Under RA, once some of the effects of a
transaction $T_i$ are observed by another transaction $T_j$,
thereafter, all effects of $T_i$ are observed by $T_j$. That is, if a
transaction $T_j$ reads a version of an object that transaction $T_i$
wrote, then a later read by $T_j$ cannot return value whose next
version is installed by $T_i$. Together with item cut isolation, RA
prevents Read Skew anomalies (Berenson et al.'s A5A) and is useful in
several contexts such as maintaining foreign key constraints,
consistent global secondary indexing, and maintenance of derived
data. In the example below, under RA, because $T_2$ has read $T_1$'s
write to $y$, $T_2$ must observe $b=c=1$ (or later versions for each
key):
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$, $a=1$, or a later version of $x$. In
the hierarchy of existing isolation properties, we place RA below
Adya's $PL-2L$ (as it does not necessarily enforce transitive
read-write dependencies) but above Read Committed ($PL-2$). Notably,
RA requires disallows reading intermediate writes (Adya's $G1b$):
observing all effects of a transaction implicitly requires observing
the final (committed) effects of the transaction as well.

Perplexingly, discussions of RA are absent from existing treatments of
weak isolation. This is perhaps again due to the single-node context
in which prior work was developed: on a single server (or a fully
replicated database), RA is achievable via lightweight locking and/or
local concurrency control over data items~\cite{gstore,
  kemme-thesis}. In contrast, in a distributed environment, RA over
arbitrary groups of non-colocated items is considerably more difficult
to achieve with high availability: servers need to ensure that,
regardless of replica selection, all of a client's accesses will read
a sufficiently up-to-date/``consistent'' set of data items. However,
replicas do not need to agree on when to choose to reveal sets of
values to clients, which would require consensus. Instead, as we see
below, RA only requires the equivalent of reliable broadcast (with
some additional client metadata) and is achievable in HAT systems.

% RA can be achieved via multi-versioning. If clients attach vector
% clocks to every write (incrementing their own position for each
% transaction) and replicas store all writes via multi-versioning, then
% clients can safely determine which sets of data items to read; this
% approach is adopted by Swift~\cite{swift} and, to a lesser-extent,
% bolt-on causal consistency~\cite{bolton}. These systems also provide
% causal consistency and are subsequently sticky available (each is
% implemented via client-side caching). However, RA is achievable
% without stickiness via an alternate algorithm: if servers wait to
% reveal writes until they are present on all replicas, clients do not
% need to be sticky.

As a straw-man, replicas can store all versions ever written for all
items. Servers can gossip information about versions they have seen
and construct a lower bound on the versions that can be found on every
replica. At the start of each transaction, clients choose a bound that
is guaranteed to be lower than or equal to the this global lower
bound, and, during transaction execution, replicas return the latest
version of each item that is not greater than the client's chosen
bound. If the lower bound is advanced along transactional boundaries,
clients will observe RA. This algorithm has several variants in the
literature~\cite{readonly, swift}: bounds can be represented by vector
clocks, and older versions can be asynchronously garbage collected.

We have developed a more efficient RA algorithm, which we sketch here
and provide greater detail in Appendix B. We begin with our Read
Committed algorithm, but replicas wait to reveal new writes to readers
until all of the replicas for the final writes in the transaction have
received their respective writes (are \textit{pending
  stable}). Clients include additional metadata with each write: a
single timestamp for all writes in the transaction (e.g., as in Read
Uncommitted) and a list of items written to in the transaction. When a
client reads, the return value's timestamp and list of items form a
lower bound on the versions that the client should read for the other
items. When a client reads, it attaches a timestamp to its request
representing the current lower bound for that item. Replicas use this
timestamp to respond with either a write matching the timestamp or a
pending stable write with a higher timestamp. Servers keep two sets of
writes for each data item: the write with the highest timestamp that
is pending stable and a set of writes that are not yet pending
stable. This is entirely master-less and operations never block due to
replica coordination. Moreover, there are several possible
optimizations that, for brevity, we do not describe here.

However, one key property with RA is that it do not dictate when
writes become visible; accordingly, the algorithm is highly
available. We can optimize write visibility via sticky availability by
sticking clients are with disjoint groups of servers: once all of a
transaction's writes are pending stable within a group of servers,
writes can be made visible (as opposed to once they are pending stable
with respect to all servers). For example, in a multi-datacenter
setting, all clients and servers within a datacenter may form a sticky
group.

@@

We illustrate the hierarchy of available, sticky available, and
unavailable consistency models we have discussed in
Figure~\ref{fig:hatcompared}. Many models are simultaneously
achievable, but we find several particularly compelling. If we combine
all sticky-HAT guarantees, we have transactional, causal snapshot
reads (i.e., Causal Transactional Predicate Cut Isolation). If we
combine RA and P-CI, we have transactional snapshot reads. We can
achieve RC, MR, and RYW by simply sticking clients to servers. We can
also combine unavailable models---for example, an unavailable system
might provide PRAM and One-Copy
Serializability~\cite{daudjee-session}.

@@

\begin{table}[t!]
\begin{tabular}{| c | p{6cm} | }\hline
HA & Read Uncommitted (RU), Read Committed (RC), Read
Atomic (RA), Item Cut Isolation (I-CI), Predicate Cut Isolation
(P-CI), Writes Follow Reads (WFR), Monotonic Reads (MR), Monotonic
Writes (MW)\\\hline Sticky HA & Read Your Writes (RYW), PRAM,
Causal\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
Snapshot Isolation (SI)\lostupdate, Repeatable Read
(RR)\lostupdate\rwskew, One-Copy Serializability
(1SR)\lostupdate\rwskew, Recency\linearizable, Safe\linearizable,
Regular\linearizable, Linearizability\linearizable, Strong
1SR\lostupdate\rwskew\linearizable \\\hline
\end{tabular}
\caption{Summary of highly available, sticky highly available, and
  unavailable models considered in this paper. Unavailable models are
  labeled by cause of unavailability: preventing lost
  update\lostupdate, preventing write skew\rwskew, and requiring
  recency guarantees\linearizable.}
\label{table:hatcompared}
\end{table}

@@

 \node[draw=none,fill=none] (ici) at (1.2, 0) {I-CI};
 \node[draw=none,fill=none] (pci) at (1.75, .95) {P-CI};
 \node[draw=none,fill=none] (rc) at (-1.2, .95) {RC};
 \node[draw=none,fill=none] (ru) at (-1.2, 0) {RU};

 \node[draw=none,fill=none] (ra) at (0, 1.475) {RA};

@@

 \node[noha](recency) at (7.7, 0) {recency};
 \node[noha](safe) at (7.7, 1) {safe};
 \node[noha](regular) at (7.7, 2) {regular};
 \node[noha](linearizable) at (7.7, 3) {linearizable};
 \node at (4.8, 2) [sticky] (causal) {causal};
 \node at (4.8, 1) [sticky] (pram) {PRAM};
 \node[noha] (cs) at (-1.2, 2) {CS};
 \node[noha] (rr) at (0.2, 2.7) {RR};
 \node[noha] (si) at (1.75, 2) {SI};
 \node[noha] (1sr) at (1.75, 3.2) {1SR};
 \node[noha] (ssr) at (3.85, 3.6) {Strong-1SR};

@@

 \draw [->] (ru) -- (rc);
 \draw [->] (rc) -- (ra);
 \draw [->] (ici) -- (pci);

@@

 \draw [->, red] (rc) -- (cs);
 \draw [->, red] (cs) -- (rr);
 \draw [->, red] (pci) -- (si);
 \draw [->, red] (ici) -- (rr);
 \draw [->, red] (rr) -- (1sr);
 \draw [->, red] (si) -- (1sr);
 \draw [->, red] (ra) -- (si);
 \draw [->, red] (ra) -- (rr);
 \draw [->, red] (causal) -- (linearizable);
 \draw [->, red] (ryw) -- (safe);

@@

\vspace{.5em}\noindent\textbf{Linearizability and Transactional
  Atomicity} The relationship between linearizability and
transactional atomicity is non-obvious. RA dictates that writes to
multiple keys across multiple servers are made visible to readers all
at once, while linearizability dictates that writes to a single key on
multiple servers are made visible to all readers at once---what is
different? First, in linearizable (and safe and regular) systems,
writes are made visible to clients \textit{immediately} after they
finish. With transactional atomicity, there is no recency
guarantee. Second, in linearizable systems, all clients see all writes
at the same time. With RA as defined here, clients may see writes at
different times depending on which replicas they contact. We are not
aware of an analogous model in the distributed systems
literature. Accordingly, despite apparent similarities, RA is
incomparable with and less coordination intensive (by availability
measures) than linearizability.

### 2013-07-22 writeup/hat-appendix.tex
Starting to address TA -> RA

\begin{definition}[Observed Transaction Vanishes (N-TA)]
A history $H$ exhibits phenomenon N-TA if $DSG(H)$ contains a directed cycle
consisting of exactly one read-dependency edge by $x$ from $T_i$ to
$T_j$ and one anti-dependency edge by $y$ from $T_j$ to $T_i$ and
$T_i$'s read from $y$ precedes its read from $x$.
\end{definition}

@@

\begin{definition}[Transactional Atomicity (TA)]
A system that provides transactional atomicity prohibits phenomenon
N-TA.
\end{definition}

@@

\begin{definition}[Snapshot Isolation]
A system that provides Snapshot Isolation prevents phenomena G0, G1a,
G1b, G1c, N-P-CI, N-TA, and Lost Update.
\end{definition}

@@

\vspace{1em}\noindent\textbf{Example.} To illustrate this algorithm consider an execution of the following transactions:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)~w_y(1)
\\T_2 &: r_x(1)~r_y(a)
\end{align*}
For TA, the system must ensure that $a=1$. Consider a system with one
replica for each of $x$ and $y$ (denoted $R_x$ and $R_y$). Say client
$c_1$ executes $T_1$ and commits. The client chooses timestamp $10001$
(whereby the last $4$ digits are reserved for client ID to ensure
uniqueness) and sends values to $w_x(1)$ to $R_x$ and $R_y$ with
$tx\_keys = \{x,y\}$ and $timestamp=10001$. $R_x$ and $R_y$ place
their respective writes into their individual \texttt{pending} sets
and send acknowledgment for transaction timestamped $10001$ to the
other. Consider the case where $r_x$ has seen $r_y$'s acknowledgment
but not vice versa: $w_x(1)$ will be in \texttt{good} on $R_x$
but \texttt{pending} on $R_y$. Now, if another client $c_2$ executes
$T_2$ and reads $x$ from $R_x$, it will read $x=1$. $c_2$ will update
its \texttt{required} vector so that $\texttt{required}=\{x:10001,
y:10001\}$. When $c_2$ reads $y$ from $R_y$, it will specify
$ts=10001$ and $R_y$ will not find a suitable write
in \texttt{good}. $R_y$ will instead return $y=1$
from \texttt{pending}.

\vspace{1em}\noindent\textbf{Overheads.} TA requires overhead both on client and server. First, clients must store a vector (\texttt{required}) for the duration of a transaction in order to determine what values to read. Second, each write contains metadata ($tx\_keys$) linear in the number of writes in the transaction. (Note that we could have stored a different timestamp with every write, but a single timestamp suffices and is more efficient) These overheads are modest for short transaction lengths but they increase value sizes on disk, in memory, and over the network. Third, on the server side, each client \textsc{put} incurs two back-ends \texttt{put}s: one into \texttt{pending} and \texttt{good}. If keeping a single value for \texttt{good}, the second \texttt{put} into \texttt{good} (as is the case with normal, single-valued eventually consistent operation) is really a \texttt{get} then a possible \texttt{put}: the server first checks if a higher-timestamped value has been written and, if not, it overwrites the current value. The \texttt{put} into \texttt{pending} is simpler: if message delivery is exactly-once, then \texttt{put}s can be made unconditionally.

>>>>

\begin{definition}[Observed Transaction Vanishes (N-RA)]
A history $H$ exhibits phenomenon N-RA if $USG(H)$ contains a directed
cycle consisting of exactly one read-dependency edge by $x$ from $T_j$
to $T_i$ and a set of edges by $y$ containing at least one
anti-dependency edge from $T_i$ to $T_j$ and $T_j$'s read from $y$
precedes its read from $x$.
\end{definition}

@@

\begin{definition}[Read Atomic (RA)]
A system that provides Read Atomic isolation prohibits phenomenon
N-RA in addition to providing Read Committed isolation.
\end{definition}

@@

\begin{definition}[Snapshot Isolation]
A system that provides Snapshot Isolation prevents phenomena G0, G1a,
G1b, G1c, N-P-CI, N-RA, and Lost Update.
\end{definition}

@@

\vspace{1em}\noindent\textbf{Example.} To illustrate this algorithm consider an execution of the following transactions:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)~w_y(1)
\\T_2 &: r_x(1)~r_y(a)
\end{align*}
For RA, the system must ensure that $a=1$. Consider a system with one
replica for each of $x$ and $y$ (denoted $R_x$ and $R_y$). Say client
$c_1$ executes $T_1$ and commits. The client chooses timestamp $10001$
(whereby the last $4$ digits are reserved for client ID to ensure
uniqueness) and sends values to $w_x(1)$ to $R_x$ and $R_y$ with
$tx\_keys = \{x,y\}$ and $timestamp=10001$. $R_x$ and $R_y$ place
their respective writes into their individual \texttt{pending} sets
and send acknowledgment for transaction timestamped $10001$ to the
other. Consider the case where $r_x$ has seen $r_y$'s acknowledgment
but not vice versa: $w_x(1)$ will be in \texttt{good} on $R_x$
but \texttt{pending} on $R_y$. Now, if another client $c_2$ executes
$T_2$ and reads $x$ from $R_x$, it will read $x=1$. $c_2$ will update
its \texttt{required} vector so that $\texttt{required}=\{x:10001,
y:10001\}$. When $c_2$ reads $y$ from $R_y$, it will specify
$ts=10001$ and $R_y$ will not find a suitable write
in \texttt{good}. $R_y$ will instead return $y=1$
from \texttt{pending}.

\vspace{1em}\noindent\textbf{Overheads.} RA requires overhead both on client and server. First, clients must store a vector (\texttt{required}) for the duration of a transaction in order to determine what values to read. Second, each write contains metadata ($tx\_keys$) linear in the number of writes in the transaction. (Note that we could have stored a different timestamp with every write, but a single timestamp suffices and is more efficient) These overheads are modest for short transaction lengths but they increase value sizes on disk, in memory, and over the network. Third, on the server side, each client \textsc{put} incurs two back-ends \texttt{put}s: one into \texttt{pending} and \texttt{good}. If keeping a single value for \texttt{good}, the second \texttt{put} into \texttt{good} (as is the case with normal, single-valued eventually consistent operation) is really a \texttt{get} then a possible \texttt{put}: the server first checks if a higher-timestamped value has been written and, if not, it overwrites the current value. The \texttt{put} into \texttt{pending} is simpler: if message delivery is exactly-once, then \texttt{put}s can be made unconditionally.
